---
title: "Police killings and the 2016 election"
author: "Eamon Caddigan, Ph.D."
date: "November 19, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
```{r libraries, include = FALSE}
library(dplyr)
library(readr)
library(stringr)
library(tidyr)
library(readxl)
library(ggplot2)
```
```{r county_data, include = FALSE}
# This code is from David Robinson: https://rpubs.com/dgrtwo/county-results
us_county_area <- read_excel("data/LND01.xls") %>%
  transmute(CountyCode = as.character(as.integer(STCOU)),
            Area = LND110210D)

us_county_population <- read_excel("data/POP01.xls") %>%
  transmute(CountyCode = as.character(as.integer(STCOU)),
            Population = POP010210D)
```
```{r police_data, include = FALSE}
# 2015 data from Mapping Police Violence: http://mappingpoliceviolence.org/aboutthedata/
police_killings <- read_csv("data/mpv_counties.csv") %>%
  transmute(CountyCode = as.character(fips)) %>%
  count(CountyCode) %>%
  rename(PoliceKillings = n)
```
```{r election_data, include = FALSE}
# More code from Robinson
election_results <- read_csv("data/election_results.csv") %>%
  group_by(CountyCode = as.character(fips)) %>%
  mutate(TotalVotes = sum(votes)) %>%
  ungroup() %>%
  mutate(name = str_replace(name, ".\\. ", "")) %>%
  filter(name %in% c("Trump", "Clinton", "Johnson", "Stein")) %>%
  transmute(County = str_replace(geo_name, " County", ""),
            State = state,
            CountyCode = as.character(fips),
            Candidate = name,
            Percent = vote_pct / 100,
            TotalVotes) %>%
  spread(Candidate, Percent, fill = 0)
```
```{r combine_data, include = FALSE}
# Combine all of the county data
county_data <- right_join(police_killings, election_results, by = "CountyCode") %>%
  replace_na(list(PoliceKillings = 0)) %>%
  left_join(us_county_population, by = "CountyCode") %>%
  left_join(us_county_area, by = "CountyCode") %>%
  mutate(KillingsPerMillionPeople = PoliceKillings / Population * 1e6,
         PopulationDensity = Population / Area)

# Select only "urban" counties (based on population density)
urban_data <- county_data %>%
  filter(PopulationDensity > 386,
         Population > 50000)
```

There's no reliable relationship between Hillary Clinton's share of the vote and the rate at which police kill people in U.S. urban counties.

```{r plot_data}
label_points <- urban_data %>%
  filter(County %in% c("District of Columbia",
                       "New York",
                       "Wyandotte",
                       "Los Angeles",
                       "Cook",
                       "Portsmouth City",
                       "Gaston",
                       "Maricopa",
                       "Harris",
                       "Gregg"))
ggplot(urban_data, 
       aes(Clinton*100, KillingsPerMillionPeople)) + 
  scale_x_continuous(breaks = seq(15, 95, by = 10)) +
  geom_point(alpha = 0.2) + 
  stat_smooth(method = "lm", fullrange = TRUE) +
  geom_text(aes(label = County), data = label_points,
            vjust = 1, hjust = 1, check_overlap = FALSE) +
  geom_point(data = label_points) +
  labs(title = "Police killings vs. Clinton vote share in US counties",
       x = "Clinton vote share (%)",
       y = "People killed by police (per million residents)") +
  theme(panel.background = element_blank(),
        panel.grid = element_blank(),
        axis.line.x = element_line(color = "black", size = 0.5),
        axis.line.y = element_line(color = "black", size = 0.5),
        axis.text = element_text(color = "black"))
```

Urban counties are defined as those with a population over 50,000 and a population density over 1,000 people/mi^2^ (280 counties). [David Robinson](http://varianceexplained.org/) described how to access county population, land area, and election data [here](https://rpubs.com/dgrtwo/county-results). Data on police killings is taken from [Mapping Police Violence](http://mappingpoliceviolence.org/) for the year 2015.

